---
title: "ST340 Lab 2: SVD & PCA"
date: "2020--21"
output: pdf_document
---
# 1: A simple singular value decomposition

(a) Generate a realization of a $4 \times 5$ Gaussian random matrix $G$.

```{r}
set.seed(5)
G = matrix(nrow = 4, ncol = 5)

for(j in 1:5){
  for(i in 1:4){
    G[i,j] = rnorm(1)
  }
}
```




(b) Look at `?svd`.
(c) Set $U$, $d$, and $V$ by using `svd`.


```{r}
svd_G = svd(G, nv = 5)
print(svd_G)

U = svd_G$u
d = svd_G$d
V = svd_G$v

L = diag(svd_G$d)
L = cbind(L, rep(0,4))  


```

\newpage

(d) Check that `G` is equal to `U%*%Sigma%*%t(V)` (to machine precision).

```{r}

SVD_calc = (U)%*%L%*%t(V)

all.equal(G, SVD_calc)     
```



(e) Plot the singular values.

```{r}
plot(x = c(1,2,3,4), y = d, main = "Singular values")

```


(f) Compute $G_2$, the 2-rank approximation of $G$, and also compute $||G-G_2||_F$.


```{r}
G_2 = d[1]*(U[,1])%*%t(V[,1]) + d[2]*(U[,2])%*%t(V[,2])

G_G_2_frobenius = sqrt(sum((G-G_2)^2))
```



(g) Does the value agree with the theory?


```{r}
all.equal(G_G_2_frobenius, sqrt(sum(d[3:4]^2)))
```


\newpage

# 2: Image compression via the singular value decomposition

```{r}
load("pictures.rdata")
source("svd.image.compression.R")
```

Take a look at `svd.image.compression.R` and understand what the code is doing.Then run `image.compression()` here to see how well we can compress our images.
\
\
I have commented on the r file that creates the functions.


```{r}

```





# 3: PCA: Crabs

(a) Load the MASS library to access the crabs data.
    ```{r}
library(MASS)
library(factoextra)
```
(b) Read `?crabs`.

```{r}
head(crabs)
```


(c) Read in the FL, RW, CL, CW, and BD measurements.
    ```{r}
Crabs <- crabs[,4:8]
Crabs.class <- factor(paste(crabs[,1],crabs[,2],sep=""))
# Creating factor that combines the species with the sex

plot(Crabs,col=Crabs.class,pch=20)
```
(d) Read `?prcomp` and use it to obtain the principal components of a centred and scaled version of `Crabs`. Call the output of `prcomp` `Crabs.pca`'.
```{r}

Crabs.pca = prcomp(Crabs, center = T, scale. = T)

```




(e) If you `plot(Crabs.pca)` it visualizes the variances associated with the components.
    ```{r}
plot(Crabs.pca)
```
(f) Plot PC2 against PC1.

```{r}
plot(((Crabs.pca)$x)[,'PC1'], ((Crabs.pca)$x)[,'PC2'], col=Crabs.class)

str(Crabs.pca)
```


(g) Read `?pairs` and use it to find a pair of components with good separation of the classes.

```{r}
pairs(Crabs.pca$x, col=Crabs.class)
```



(h) Read `?scale`. Check that you can obtain the principal components by using the singular value decomposition on a centred and scaled version of `Crabs`.



# 4: PCA: Viruses

This is a dataset on 61 viruses with rod-shaped particles affecting various crops (tobacco, tomato, cucumber and others) described by Fauquet *et al.* (1988) and analysed by Eslava-G$\'o$mez (1989). There are 18 measurements on each virus, the number of amino acid residues per molecule of coat protein.
```{r}
load("viruses.rdata")
```

(a) Obtain the principal components of a centred and scaled version of allviruses.
    ```{r}
groups <- rep(0,61)
groups[1:3] <- 1
groups[4:9] <- 2
groups[10:48] <- 3
groups[49:61] <- 4
group.names <- c("Hordeviruses","Tobraviruses","Tobamoviruses","furoviruses")

head(allviruses)

allviruses.PCA = prcomp(allviruses, center = T, scale. = T)

```
If you colour by groups (i.e. `col=groups` in plot) then black is horde, red is tobra, green is tobamo, blue is furo.

(b) Do the principal components show some separation between the viruses?



```{r}

fviz_eig(allviruses.PCA)

pairs(allviruses.PCA$x[,1:5], col = groups)
```




(c) The largest group of viruses is the tobamoviruses. Does a principal component analysis suggest there might be subgroups within this group of viruses?

```{r}




tobamoviruses.PCA = prcomp(tobamoviruses, center = T, scale. = T)

fviz_eig(tobamoviruses.PCA)

plot(tobamoviruses.PCA$x[,1], tobamoviruses.PCA$x[,2])
pairs(tobamoviruses.PCA$x[,1:3])

```


From the plot it looks like we have 3 clusters, meaning that there are maybe groups that have different characteristics within the Tobamoviruses group of viruses.



