---
title: "ST340 Lab 3: Markov chains & PageRank, K-means"
date: "2020--21"
output: pdf_document
---
# 1: PageRank



Inspect and run the following code, which sets up each object we need in PageRank.

* `links[,1]` denotes "from" page
* `links[,2]` denotes "to" page

Note: there are no links from a page to itself.  `A.sparse` is a sparse matrix with `A[i,j] = 1` whenever `links[k,]` $= (i,j)$ for some `k`.
\
This links dataset gives links between pages. Where the the first column denotes where the link starts and the second column tell us the page where the link goes.


```{r, cache=TRUE}
library(Matrix) # Load the sparse Matrix library
load("web-google.rdata") # Load the Google web data:


n <- max(links)             # Equal to the number of pages
numlinks <- dim(links)[1]   # number of links



# Turn the data into an adjacency matrix:
A.sparse <- sparseMatrix(i=links[,1], j=links[,2], x=rep(1,numlinks), dims=c(n,n))         # This the matrix A in the notes
```

A.sparse having entry 1 at element [i,j], means there exists a link from page i to page j.

```{r, cache=TRUE}
outlinks <- rep(0,n)
for (i in 1:numlinks) {
  # Calculate outlinks where `outlinks[i]` is the number of outlinks for page `i`:
  outlinks[links[i,1]] <- outlinks[links[i,1]] + 1
}

d <- outlinks == 0         # Calculate `d`: the binary vector that is 1 when page `i` has no outlinks





vsH <- rep(0,numlinks)
for (k in 1:numlinks) {
  # Calculate the values to assign to `H`. These are nonzero whenever links[k,]` = (i,j)
  # for some k but are normalized so that sum(A[i,]) = 1 whenever sum(A[i,]) > 0
  # (i.e. apart from when i is a dangling node).
  vsH[k] <- 1/outlinks[links[k,1]]        # this gives 1/(the number outlinks)
}

# Construct H as a sparse matrix
H.sparse <- sparseMatrix(i=links[,1], j=links[,2], x=vsH, dims=c(n,n)) 

alpha <- .85               # damping factor
w <- rep(1/n,n)            # (these are not sensible choices of w and p, but they "work")
p <- log(1+(1:n))/sum(log(1+(1:n)))          # personalization probability
```







## 2: Power Iteration

(a) We will do power iteration for `m` iterations:
```{r}
m <- 150
```
(b) We will compute the difference between probability vectors after each iteration:
```{r}
diffs <- rep(0,m)
```

(c) Start with the uniform distribution:
```{r eval=FALSE}
muT <- t(rep(1/n,n))
for (i in 1:m) {
  muT.old <- muT # store the old value of muT
  
  # compute the new value of muT = muT.old%*%G
  
  ## write your code here
  muT = (alpha*((muT.old)%*%(H.sparse))) + (alpha*(muT.old%*%d)%*%(t(w))) + (1 - alpha)%*%t(p)
  
  # compute the difference
  diff <- sum(abs(muT-muT.old))
  diffs[i] <- diff
  print(paste("iteration ",i,": ||muT-muT.old||_1 = ",diff,sep="",sum(muT)))
}
```

(d) Rank the pages!
```{r eval=FALSE}
ranking <- sort(muT, decreasing=TRUE, index.return=TRUE)$ix
```







## 3: K-means

(a) Write a $k$-means algorithm:
```{r }
my.kmeans <- function(xs,K,ZZZZzzzz=0) {
  
  xs = as.matrix(xs)
  
  # find the number of columns and rows of xs
  p <- dim(xs)[2]
  n <- dim(xs)[1]
  
  # initialize the clusters
  minx <- min(xs)
  maxx <- max(xs)
  cs <- xs[sample(n,K),]    # Chosing K random centres
  
  # zs will be modified in the code
  zs <- rep(0,n)
  
  
  converged <- FALSE
  while(!converged) {
    cs.old <- cs
    zs.old <- zs
    
    # update zs
    dist_x_c = matrix(nrow=n, ncol=K)
    for(i in 1:n){
      for(j in 1:K){
        dist_x_c[i,j] = sum(((xs[i,]) - (cs[j,]))^2)
      }
    }
    
    dist_x_c_min = rep(0,n)
    for(i in 1:n){
      zs[i] = (1:K)[dist_x_c[i,] == min(dist_x_c[i,])]
    }
    
    
    plot(xs,col=zs,pch=20)
    points(cs,pch=20,col="blue",cex=2)
    
    Sys.sleep(ZZZZzzzz)
    
    # update cs
    
    for(i in 1:K){
      cs[i,] = cbind(mean(xs[zs == i,1]), mean(xs[zs == i,2]))
    }
    
    plot(xs,col=zs,pch=20)
    points(cs,pch=20,col="blue",cex=2)
    
    Sys.sleep(ZZZZzzzz)
    
    if (all(zs==zs.old)){
      converged <- TRUE
    }
  }
  return(list(cs=cs,zs=zs))
}
```
(b) Check that it works:
```{r }
xs <- faithful
my.kmeans.out <- my.kmeans(xs, K=2, ZZZZzzzz=1)

```

(c) Check the output against the output of `kmeans(faithful,2)`:
```{r }
par(mfrow=c(1,2))
plot(xs, col=my.kmeans.out$zs, pch=20)
points(my.kmeans.out$cs ,pch=20, col="blue", cex=2)

builtin.kmeans.out <- kmeans(xs,2)
plot(xs,col=builtin.kmeans.out$cluster,pch=20)
points(builtin.kmeans.out$centers,pch=20,col="blue",cex=2)
```

```{r echo=FALSE, fig.cap="Built-in `kmeans` function for the Old Faithful datasset."}
xs <- faithful
builtin.kmeans.out <- kmeans(xs,2)
plot(xs, col=builtin.kmeans.out$cluster, pch=20)
points(builtin.kmeans.out$centers,pch=20,col="blue",cex=2)
```

